REVISITING SEMI-CONTINUOUS HIDDEN MARKOV MODELS

### 2011-09-27 misc/papers/icassp12_tied/icassp12.tex
updates to semicont paper
git-svn-id: https://svn.code.sf.net/p/kaldi/code/sandbox/discrim@526 5e6a8d80-dfce-4ca6-a32a-6e07a63d50c8

\begin{abstract}
In the past decade, semi-continuous hidden Markov models (SC-HMMs) have not 
attracted much attention in the speech recognition community. Growing amounts 
of training data and increasing sophistication of model estimation led to the 
impression that continuous HMMs are the best choice of acoustic model.
%
However, recent work on recognition of under-resourced languages faces the same
old problem of estimating a large number of parameters from limited amounts 
of transcribed speech.
This has led to a renewed interest in methods of reducing the number of parameters 
while maintaining or extending the modeling capabilities of continuous models.
%
In this work, we compare continuous, classic and multiple-codebook semi-continuous,
with full covariance matrices, and subspace Gaussian mixture models.
%
Experiments using on the RM and WSJ corpora show that a semi-continuous system
can still yield competitive results while using fewer Gaussian components.
\end{abstract}

@@

\subsection{Acoustic-Phonetic Decision Tree Building}
The acoustic-phonetic decision tree provides the link between phones
in context and emission probability density functions (pdfs). 
%
The statistics for the tree clustering consist of the 
sufficient statistics needed to estimate a Gaussian for each seen combination 
of (phonetic context, HMM-state).  These statistics are based on a Viterbi 
alignment of a previously built system.  The roots
of the decision can be configured; for the RM experiments, these correspond to
the phones in the system, and on Wall Street Journal, they
correspond to the ``real phones'', i.e. grouping different stress and
position-marked versions of the same phone together.

@@

\subsection{Two-level Tree based Multiple-Codebook Semi-continuous Models}
A style of system that lies between the semi-continuous and continuous types of
systems is one based on a two-level tree; this is described in~\cite{prasad2004t2b}
as a state-clustered tied-mixture (SCTM) system. The way we implement this is
to first build the decision tree used for phonetic clustering to a relatively
coarse level (e.g.~100); and to then split more finely (e.g.~2500), but for
each new leaf, remember which leaf of the original tree it corresponds to.
Each leaf of the first tree corresponds to a codebook of Gaussians, i.e.~the 
Gaussian parameters are tied at this level.  For this type of system we do not
apply any post-clustering.
The leaves of the second tree contain the weights, and these leaves correspond 
to the actual context-dependent HMM states.

@@

\section{Smoothing Techniques for Semi-continuous Models}
%
\subsection{Intra-Iteration Smoothing}
Although an increasing number of tree leaves does not imply a huge increase in
parameters, the sufficient statistics for the individual leaves will be
sparser, especially for little training data.
The {\em intra-iteration} smoothing is motivated by the fact that tree leaves
that share the root are similar, thus, if the sufficient statistics of the
weights of a leaf $j$ are very small, they should be interpolated with the 
statistics of closely related leaf nodes.
To do so, we first propagate all sufficient statistics up the tree so that the
statistics of any node is the sum of its children's statistics.
Second, the statistics of each node and leaf are interpolated top-down with 
their parent's using an interpolation weight $\rho$ in
\begin{equation} \label{eq:intra}
\hat\gamma_{ji} \leftarrow 
  \underbrace{\left(\gamma_{ji} + \frac{\rho}{\left( \sum_k \gamma_{p(j),k} \right) + \epsilon} \, \gamma_{p(j),i}\right)}_{\mathrel{\mathop{:}}= \bar\gamma_{ji}}
  \cdot 
  \underbrace{\frac{\sum_k \gamma_{jk}}{\sum_k \bar\gamma_{jk}}}_\text{normalization}
\end{equation}
where $\gamma_{ji}$ is the accumulator of weight $i$ in tree leaf $j$
and $p(j)$ is the mapping to the parent node of $j$.
%
For two-level tree based models, the propagation and interpolation cannot
be applied across different codebooks. 
%
A similar yet different interpolation scheme without normalization was used in 
\cite{schukattalamazzini1994srf,schukattalamazzini1995as} along with a different
tree structure.

@@

\subsection{Inter-Iteration Smoothing}
Another typical problem that arises when training semi-continuous models is
that the weights tend to converge on a local optimum over the iterations. 
This is due to the fact that the codebook parameters change rather slowly 
as the statistics are collected from all respective tree leaves, but the 
leaf specific weights converge much faster.
%
To compensate for this, we smooth the newly estimated {\em parameters} 
$\Theta^{(i+1)}$ with the ones from the prior iteration using an interpolation
weight $\varrho$
\begin{equation}
\hat\Theta^{(i+1)} \leftarrow (1 - \varrho) \, \Theta^{(i+1)} + \varrho \, \Theta^{(i)} % \quad .
\end{equation}
which leads to an exponentially decreasing impact of the initial parameter set.

@@

\subsection{Resource Management}
For the RM data set, we train on the speaker independent training and development 
set (about 4000 utterances) and test on the six DARPA test runs Mar and Oct '87,
Feb and Oct '89, Feb '91 and Sep '92 (about 1500 utterances total).  The parameters 
were tuned to the test set.
%
\begin{itemize}
\item
  {\em cont}: continuous triphone system using 9011 diagonal covariance
  Gaussians in 1480 tree leaves
\item
  {\em semi}: semi-continuous triphone system using 768 full covariance
  Gaussians in 2500 tree leaves, no smoothing (explanations later).
\item
  {\em 2lvl}: two-level tree based semi-continuous triphone system using
  3072 full covariance Gaussians in 208 codebooks ($4/14.7/141$ min/average/max
  components), 2500 tree leaves, $\rho = 35$ and $\varrho = 0.2$.
\item
  {\em sgmm}: subspace Gaussian mixture model triphone system using a 400 
  full-covariance Gaussian background model and 2016 tree leaves.
\end{itemize}

\subsection{WSJ}
% TODO fix the WSJ train/dev/eval description
%% Korbinian: For now, I inserted results on si84/half, using cmvn (tri2c),
%             the si284 take just too long for the submission.
%For the WSJ data set, we trained on the SI-284 training set, tuned on dev
%and tested on eval92,93.
To get an intuition of the performance on the WSJ data set, we trained on 
half of the SI-84 training set using settings similar to the RM experiments; 
we tested on the eval '92 and eval '93 data sets.
As the classic semi-continuous system did not yield good performance on the
RM data both in terms of run-time and performance, we omit it for the WSJ
experiments.
%
% TODO Korbinian: I hope the SI-284 experiments finish soon enough while the 
%                 submission is still open. For now, I describe SI-84/half.
\begin{itemize}
\item
  {\em cont}: continuous triphone system using 10000 diagonal covariance
  Gaussians in 1576 tree leaves.
%% TODO Korbinian: For now, I use the same setting as in RM, to evade the
%                  tuning on the dev set. Also, no semi system, too slow
%                  and bad performance on RM.
\item
    {\em 2lvl}: same configuration as for RM (no further tuning).
%\item
%  {\em 2lvl}: two-level tree based semi-continuous triphone system using
%  4096 full covariance Gaussians in 208 codebooks ( min/average/max
%  components),  tree leaves, $\rho = $ and $\varrho = $.
\item
  {\em sgmm}: subspace Gaussian mixture model triphone system using a 400 
  full-covariance Gaussian background model and 2361 tree leaves.
\end{itemize}

@@

The results on the different test sets on the RM data are displayed in 
Tab.~\ref{tab:res_rm}. The classic semi-continuous system shows the worst 
performance of the four. Possible explanations are an insufficient number
of Gaussians paired with overfitting of the weights to a little number of 
components. Unfortunately, this could not be eased with the interpolations
and larger numbers of Gaussians would unacceptably increase the computational 
load due to the full covariance matrices.

@@

\begin{table}%[tb]
\begin{center}
\begin{tabular}{|c|c||c|c|c|c|}
\hline
covariance & Gaussians & none & inter & intra & both \\ \hline\hline
full       &      1024 & 3.70 & 3.64 & 3.79 & 3.74 \\ \hline
full       &      3072 & 3.01 & 3.01 & 3.02 & 2.90 \\ \hline\hline
diagonal   &      3072 & 4.13 & 4.15 & 4.25 & 4.35 \\ \hline
diagonal   &      9216 & 3.22 & 3.09 & 3.28 & 3.20 \\ \hline
\end{tabular}
\end{center}
\caption{\label{tab:rm_diagfull}
Average \% WER of the multiple-codebook semi-continuous model using different 
numbers diagonal and full covariance Gaussians, and different smoothings 
on the RM data. 
Settings are 208 codebooks, 2500 context dependent states; $\rho = 35$ 
and $\varrho = 0.2$ if active.
}
\end{table}

Keeping the number of codebooks and leaves as well as the smoothing parameters
$\rho$ and $\varrho$ of {\em 2lvl} constant, we experiment with the number of
Gaussians and type of covariance; the results are displayed in 
Tab.~\ref{tab:rm_diagfull}.
Interestingly, the full covariances make a strong difference: Using the same
number of Gaussians, full covariances lead to a significant improvement.
On the other hand, a substantial increase of the number of diagonal Gaussians 
leads to a similar performance as the regular continuous system.
%
Another observation from Tab.~\ref{tab:rm_diagfull} is that the smoothing
parameters need to be carefully calibrated. While the inter-iteration smoothing
helps in most cases, the intra-iteration smoothing coefficient $\rho$ strongly
depends on the type and number of Gaussians, which is due to the direct influence
of $\rho$ on the counts in Eq.~\ref{eq:intra}.

@@

The results on the different test sets on the WSJ data are displayed in
Tab.~\ref{tab:res_rm}.
% TODO Korbinian: The following is for the si-84/half results, as currently
%                 in the table.
Using the parameters calibrated on the RM test data, the {\em 2lvl} system
performance is not as good as of the continuous or GMM systems, but still
in a similar range while using about the same number of leaves but about a
third of the Gaussians.
Once the number of Gaussians and leaves, as well as the smoothing parameters 
are tuned on the development set, we expect the error rates to be in between 
the continuous and SGMM systems, as seen on the RM data.

>>>>

\begin{abstract}
In the past decade, semi-continuous hidden Markov models (SC-HMMs) have not 
attracted much attention in the speech recognition community. Growing amounts 
of training data and increasing sophistication of model estimation led to the 
impression that continuous HMMs are the best choice of acoustic model.
%
However, recent work on recognition of under-resourced languages faces the same
old problem of estimating a large number of parameters from limited amounts 
of transcribed speech.
This has led to a renewed interest in methods of reducing the number of parameters 
while maintaining or extending the modeling capabilities of continuous models.
%
In this work, we compare classic and multiple-codebook semi-continuous models 
that use full covariance matrices with continuous HMMs and subspace Gaussian mixture models.
%
Experiments using on the RM and WSJ corpora show that a semi-continuous system
can still yield competitive results while using fewer Gaussian components.
\end{abstract}

@@

\subsection{Acoustic-Phonetic Decision Tree Building}
The acoustic-phonetic decision tree provides a mapping between phones
in context and emission probability density functions (pdfs). 
%
The statistics for the tree clustering consist of the 
sufficient statistics needed to estimate a Gaussian for each seen combination 
of (phonetic context, HMM-state).  These statistics are based on a Viterbi 
alignment of a previously built system.  The roots
of the decision can be configured; for the RM experiments, these correspond to
the phones in the system, and on Wall Street Journal, they
correspond to the ``real phones'', i.e. grouping different stress and
position-marked versions of the same phone together.

@@

\subsection{Multiple-Codebook Semi-continuous Models}
A style of system that lies between the semi-continuous and continuous types of
systems is one based on a two-level tree; this is described in~\cite{prasad2004t2b}
as a state-clustered tied-mixture (SCTM) system. The way we implement this is
to first build the decision tree used for phonetic clustering to a relatively
coarse level (e.g.~100 leaves). Each leaf of this tree corresponds to a codebook of Gaussians, i.e.~the Gaussian parameters are tied at this level. The leaves 
of this tree are then split further to give a more fine-grained clustering (e.g.~2500 leaves), where for each new leaf we remember which codebook it corresponds to. The leaves of this second tree correspond to the actual context-dependent HMM states and contain the weights.
For this type of system we do not apply any post-clustering.

@@

\section{Smoothing Techniques for Semi-continuous Models}
%
\subsection{Intra-Iteration Smoothing}
Although an increasing number of tree leaves does not imply a huge increase in
parameters, the estimation of weights at the individual leaves suffer from data 
fragmentation, especially when training models with small amounts of data.
The {\em intra-iteration} smoothing is motivated by the fact that tree leaves
that share the root are similar, thus, if the sufficient statistics of the
weights of a leaf $j$ are very small, they should be interpolated with the 
statistics of closely related leaf nodes.
To do so, we first propagate all sufficient statistics up the tree so that the
statistics of any node is the sum of its children's statistics.
Second, the statistics of each node and leaf are interpolated top-down with 
their parent's statistics using an interpolation weight $\tau$:
\begin{equation} \label{eq:intra}
\hat\gamma_{ji} \leftarrow 
  \underbrace{\left(\gamma_{ji} + \frac{\tau}{\left( \sum_k \gamma_{p(j),k} \right) + \epsilon} \, \gamma_{p(j),i}\right)}_{\mathrel{\mathop{:}}= \bar\gamma_{ji}}
  \cdot 
  \underbrace{\frac{\sum_k \gamma_{jk}}{\sum_k \bar\gamma_{jk}}}_\text{normalization},
\end{equation}
where $\gamma_{ji}$ is the occupancy of the $i$ component in tree leaf $j$
and $p(j)$ is the parent node of $j$.
%
For two-level tree based models, the propagation and interpolation cannot
be applied across different codebooks. 
%
A similar yet different interpolation scheme without normalization was used in 
\cite{schukattalamazzini1994srf,schukattalamazzini1995as} along with a different
tree structure.

@@

\subsection{Inter-Iteration Smoothing}
Another typical problem that arises when training semi-continuous models is
that the weights tend to converge to a local optimum over the iterations. 
This is due to the fact that the codebook parameters change rather slowly 
as the statistics are collected from all respective tree leaves, but the 
leaf specific weights converge much faster.
%
To compensate for this, we smooth the newly estimated parameters $\Theta^{(i+1)}$ with the ones from the prior iteration using an interpolation weight $\rho$
\begin{equation}
\hat\Theta^{(i+1)} \leftarrow (1 - \rho) \, \Theta^{(i+1)} + \rho \, \Theta^{(i)} % \quad .
\end{equation}
which leads to an exponentially decreasing impact of the initial parameter set.

@@

\subsection{Resource Management}
For the RM data set, we train on the speaker independent training and development 
set (about 4000 utterances) and test on the six DARPA test runs Mar and Oct '87,
Feb and Oct '89, Feb '91 and Sep '92 (about 1500 utterances total).  The parameters 
are tuned to the test set since RM has no hold-out data set. We use the following system identifiers in Table~\ref{tab:res_rm}:
%
\begin{itemize}
\item
  {\em cont}: continuous triphone system using 9011 diagonal covariance
  Gaussians in 1480 tree leaves
\item
  {\em semi}: semi-continuous triphone system using 768 full covariance
  Gaussians in 2500 tree leaves, no smoothing (explanations later).
\item
  {\em 2lvl}: two-level tree based semi-continuous triphone system using
  3072 full covariance Gaussians in 208 codebooks ($4/14.7/141$ min/average/max
  components), 2500 tree leaves, $\tau = 35$ and $\rho = 0.2$.
\item
  {\em sgmm}: subspace Gaussian mixture model triphone system using a 400 
  full-covariance Gaussian background model and 2016 tree leaves.
\end{itemize}

\subsection{WSJ}
% TODO fix the WSJ train/dev/eval description
%% Korbinian: For now, I inserted results on si84/half, using cmvn (tri2c),
%             the si284 take just too long for the submission.
%For the WSJ data set, we trained on the SI-284 training set, tuned on dev
%and tested on eval92,93.
To get an intuition of the performance on the WSJ data set, we trained on 
half of the SI-84 training set using settings similar to the RM experiments; 
we tested on the eval '92 and eval '93 data sets.
As the classic semi-continuous system did not yield good performance on the
RM data both in terms of run-time and performance, we omit it for the WSJ
experiments. Results corresponding to the following systems are presented in 
Table~\ref{tab:res_wsj}:
%
% TODO Korbinian: I hope the SI-284 experiments finish soon enough while the 
%                 submission is still open. For now, I describe SI-84/half.
\begin{itemize}
\item
  {\em cont}: continuous triphone system using 10000 diagonal covariance
  Gaussians in 1576 tree leaves.
%% TODO Korbinian: For now, I use the same setting as in RM, to evade the
%                  tuning on the dev set. Also, no semi system, too slow
%                  and bad performance on RM.
\item
    {\em 2lvl}: same configuration as for RM (no further tuning).
%\item
%  {\em 2lvl}: two-level tree based semi-continuous triphone system using
%  4096 full covariance Gaussians in 208 codebooks ( min/average/max
%  components),  tree leaves, $\tau = $ and $\rho = $.
\item
  {\em sgmm}: subspace Gaussian mixture model triphone system using a 400 
  full-covariance Gaussian background model and 2361 tree leaves.
\end{itemize}

@@

The results on the different test sets on the RM data are displayed in 
Table~\ref{tab:res_rm}. The classic semi-continuous system shows the worst 
performance of the four. Possible explanations are an insufficient number
of Gaussians paired with overfitting of the weights to a little number of 
components. Unfortunately, this could not be eased with the interpolations
and larger numbers of Gaussians would unacceptably increase the computational 
load due to the full covariance matrices.

@@

\begin{table}%[tb]
\begin{center}
\begin{tabular}{|c|c||c|c|c|c|}
\hline
covariance & Gaussians & none & inter & intra & both \\ \hline\hline
full       &      1024 & 3.70 & 3.64 & 3.79 & 3.74 \\ \hline
full       &      3072 & 3.01 & 3.01 & 3.02 & 2.90 \\ \hline\hline
diagonal   &      3072 & 4.13 & 4.15 & 4.25 & 4.35 \\ \hline
diagonal   &      9216 & 3.22 & 3.09 & 3.28 & 3.20 \\ \hline
\end{tabular}
\end{center}
\caption{\label{tab:rm_diagfull}
Average \% WER of the multiple-codebook semi-continuous model using different 
numbers diagonal and full covariance Gaussians, and different smoothings 
on the RM data. 
Settings are 208 codebooks, 2500 context dependent states; $\tau = 35$ 
and $\rho = 0.2$ if active.
}
\end{table}

Keeping the number of codebooks and leaves as well as the smoothing parameters
$\tau$ and $\rho$ of {\em 2lvl} constant, we experiment with the number of
Gaussians and type of covariance; the results are displayed in 
Table~\ref{tab:rm_diagfull}.
Interestingly, the full covariances make a strong difference: Using the same
number of Gaussians, full covariances lead to a significant improvement.
On the other hand, a substantial increase of the number of diagonal Gaussians 
leads to a similar performance as the regular continuous system.
%
Another observation from Table~\ref{tab:rm_diagfull} is that the smoothing
parameters need to be carefully calibrated. While the inter-iteration smoothing
helps in most cases, the intra-iteration smoothing coefficient $\tau$ strongly
depends on the type and number of Gaussians, which is due to the direct influence
of $\tau$ on the counts in equation~\ref{eq:intra}.

@@

The results on the different test sets on the WSJ data are displayed in
Table~\ref{tab:res_rm}.
% TODO Korbinian: The following is for the si-84/half results, as currently
%                 in the table.
Using the parameters calibrated on the RM test data, the {\em 2lvl} system
performance is not as good as of the continuous or GMM systems, but still
in a similar range while using about the same number of leaves but about a
third of the Gaussians.
Once the number of Gaussians and leaves, as well as the smoothing parameters 
are tuned on the development set, we expect the error rates to be in between 
the continuous and SGMM systems, as seen on the RM data.
